Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks

### 2025-05-14 inst/paper/main.tex
update methods with H7N9 info, tidy end of methods, correctly cite all references in methods, fix wording in results and reorder presymptomatic and asymptomatic results

The time between an infector being infected and then infecting an infectee (i.e. generation time) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector. The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 - 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution). In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector. \\

@@

There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks. We set this limit to 500 cases. This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e. larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).

@@

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}. H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}. \\

@@

To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment. We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmissin, and number of initial cases. \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/onset_to_isolation.png}
\caption{Three assumed onset-to-isolation delays used in the simuluation experiment. The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong (Donnelly et al. 2003), the median onset-to-isolation time is 3.4 days (blue dashed line). The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China (Li et al., 2020), the median onset-to-isolatiojn is 8.1 days (red dashed line). The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days.}
\label{fig:onset-to-isolation}
\end{figure}

@@

We define the proportion of pre-symptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$). The shape parameter controls the proportion of pre-symptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}). We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:prop-presym-trans}). Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective contact tracing is hypothesied to be. \\

The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks. There is evidence from past human outbreaks of H1N1pdm and H5N1 (2.3.4.4b) that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009, gargHighlyPathogenicAvian2025}; but whether human-to-human transmission can occur while subclinical is unknown. There is no evidence of asymptomatic transmission for the H7N9 subtype \citep{xuSerologicalInvestigationSubclinical2013}. Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.  We ran three scenarios, one with zero asymptomatic tranmission, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing. \\

@@

We simulated scenarios with either 5, 20 or 40 initial cases. This is to test whether contact tracing can contain outbreaks that may have already accumulated cases before contact tracing is started, or represents importation of multiple cases in a single event, such as a flight with multiple infectors (Khanh et al. 2020). \\

The proportion of ascertainment is varied between 0 and 1, at intervals of 0.2, to test at what proportion of missed cases would lead to outbreaks becoming uncontrolled epidemics. \\

Some parameters we fixed across all simulations. The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, 0.16 for subclinical and community cases. For computational efficiency we capped the maximum number of days the simulation could run for to 365. This is substantially longer than the 12-16 week window for evaluating outbreak control so does not influence the results. Additionally, we limit simulations to 500 cases. This is an arbitrary threshold we judge to be large epidemic that is not controllable. \\

@@

Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}). A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of contacts traced, except in cases where they are equal because zero or all outbreaks are contained, across all onset-to-isolation scenarios (Figure \ref{fig:prop-outbreak-control-R}). When the reponse to isolate cases is slow then outbreaks with $R = 1.5$ require at least 50\% of contacts traced to contain at least half of simulated outbreaks. When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$ to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A). Especially for H1N1, given it's faster tempo of transmission, having a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A). In a scenario with a fast, Hong Kong SARS-like, onset-to-isolation response, outbreaks with $R = 1.1$ are almost always contained at any proportion of contacts traced (Figure \ref{fig:prop-outbreak-control-R}B). For slow and fast isolation response delays there is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}A,B). As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e. 0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B). For more transmissible influenza scenarios, even a fast isolation response with an average of 3.4 days between symptom onset and isolation requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B). \\

The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}). The contol of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C). Assuming ubiquitous access to home testing and that 63\% of symptomatic cases testing and isolating within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$). The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C). \\

The difference in influenza subtype incubation period influences the efficacy of isolating and tracing. This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e. not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}). The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission. The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e. difference between circle, triangle and square points at each contacts traced \% in Figure \ref{fig:prop-outbreak-control-R}). This variance is greatest at intermediate (20\% - 80\%) contact tracing effectiveness and intermediate values of $R$ (1.5 and 2.5) (Figure \ref{fig:prop-outbreak-control-R}). This could indicate a substantive role for incubation period durartion in the control of disease transmission. However, this variance in proportion of outbreaks controlled between pathogen subtypes does not show a strong pattern across all scenarios simulated (Figure \ref{fig:prop-outbreak-control-var-R}). There is a weak pattern of increased variance in controllability between subtypes at intermediate values of $R$, overall it seems that incubation period causes differences in control-by-isolation effectivness when not at the boundary of epidemic controllability. In other words, differences between influenza subtypes arise when not in a situation of high transmissibility and low contact tracing effectivness that cannot contain any outbreaks, nor low transmissibility and high contact tracing effectivenss easily contain all outbreaks (Figure \ref{fig:prop-outbreak-control-var-R}). \\

@@

The pattern of outbreak control is broadly similar between 10\% asymptomatic cases and the 0\% and 30\% asymptomatic scenarios (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). Qualitative difference are in the 30\% asymptomatic scenario for LFT isolation, where the ability to control highly transmissible influenzas reduces to less than a quarter of outbreaks controlled at maxmimum contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). When every case is symptomatic the likelihood of containing highly transmissible outbreaks drastically increases for all onset-to-isolation response delays (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). \\

@@

The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak. The higher the number of cases the less chance that stochastic extinction of a transmission chain will end the whole epidemic. For an influenza pandemic, the number of initial infected individuals influences efficacy of control-by-isolation at all transmissibility scenarios evaluated ($R \in \{1.1, 1.5, 2.5, 3.5\}$, Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.1$ most outbreaks can be controlled without isolation and contact tracing, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}). When the number of cases seeding the outbreak is $>20$ at $R>2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $>60\%$. In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks (Figure \ref{fig:prop-outbreak-control-num-init-cases}).  \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_onset_to_isolation.png}
\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an onset-to-isolation delay distribution that is either SARS-like, or COVID-19-like (from Wuhan). This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the number of initial infections is 10.}
\label{fig:prop-outbreak-control-onset-to-isolation}
\end{figure}

>>>>

The time between an infector being infected and then infecting an infectee (i.e. generation time) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector. The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution). In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector. \\

@@

There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks. We set this limit to 500 cases. This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e. larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).

@@

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previously caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}. H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}. The first human case with the H7N9 subtype was in 2013, subsequently causing five outbreak waves, but no human cases since 2019 \citep{liH7N9InfluenzaVirus2021}. Most cases are of zoonotic origin with recorded animal exposure but there have been some small clusters (2 --- 3 cases) of possible human-to-human transmission \citep{liEpidemiologyHumanInfections2014}. For the purposes of this study we assume that there is no preexisting population immunity for these influenza subtypes. This is generally true for H5N1 and H7N9 at present, and for H1N1pdm09 prior to 2009. \\

@@

To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment. We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmission, and number of initial cases. \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/onset_to_isolation.png}
\caption{Three assumed onset-to-isolation delays used in the simuluation experiment. The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolatiojn is 8.1 days (blue dashed line). The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line). The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).}
\label{fig:onset-to-isolation}
\end{figure}

@@

We define the proportion of presymptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$). The shape parameter controls the proportion of presymptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}). We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:prop-presym-trans}). Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective isolation and contact tracing is hypothesied to be. \\

The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks. There is evidence from past human outbreaks of H1N1pdm and H5N1 (2.3.4.4b) that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009, gargHighlyPathogenicAvian2025}; but whether human-to-human transmission can occur while subclinical is unknown. There is no evidence of asymptomatic transmission for the H7N9 subtype \citep{xuSerologicalInvestigationSubclinical2013}. Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.  We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing. \\

@@

The feasibility of epidemic containment by contact tracing is expected to depend on the proportion of close contacts of cases that are traced and isolated if symptomatic. We varied the proportion of contacts ascertained in the model between 0 and 1, at intervals of 0.2. In cases when the proportion of contacts traced is zero, isolation after the delay from symptom onset is the only NPI reducing disease transmission. To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial indepedent cases. This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases. \\

We fixed some parameters across all simulations. The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, and 0.16 for subclinical and community cases. For computational efficiency we capped the maximum number of days the simulation could run for to 365. This is substantially longer than the 12-16 week window for evaluating outbreak control so does not influence the results. Additionally, we limit simulations to 500 cases. This is an arbitrary threshold we judge to be large epidemic that is not controllable.

@@

Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}). A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of contacts traced, except in cases where they are equal because zero or all outbreaks are contained, across all onset-to-isolation scenarios (Figure \ref{fig:prop-outbreak-control-R}). When the reponse to isolate cases is slow then outbreaks with $R = 1.5$ require at least 50\% of contacts traced to contain at least half of simulated outbreaks. When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$) to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A). Especially for H1N1, given its faster tempo of transmission, has a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A). In a scenario with a fast, Hong Kong SARS-like, onset-to-isolation response, outbreaks with $R = 1.1$ are almost always contained at any proportion of contacts traced (Figure \ref{fig:prop-outbreak-control-R}B). For slow and fast isolation response delays there is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}A,B). As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e. 0\% contact ascertainment), however, outbreaks with $R = 1.5$ are mostly contained once the percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B). For more transmissible influenza scenarios, even a fast isolation response, with an average of 3.4 days between symptom onset and isolation, requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B). \\

The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}). The contol of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C). Assuming ubiquitous access to home testing and that 63\% of symptomatic cases test and isolate within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$). The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C). \\

The difference in influenza subtype incubation period influences the efficacy of isolating and tracing. This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e. not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}). The shortest incubation period, H1N1, is predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively trace and isolate their potential infectees in time to prevent disease transmission. The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e. difference between circle, triangle and square points at each contacts traced \% in Figure \ref{fig:prop-outbreak-control-R}). This variance is greatest at intermediate (20\% - 80\%) contact tracing effectiveness and intermediate values of $R$ (1.5 and 2.5) (Figure \ref{fig:prop-outbreak-control-R}). This could indicate a substantive role for incubation period durartion in the control of disease transmission. However, this variance in proportion of outbreaks controlled between pathogen subtypes does not show a strong pattern across all scenarios simulated (Figure \ref{fig:prop-outbreak-control-var-R}). There is a weak pattern of increased variance in controllability between subtypes at intermediate values of $R$, overall it seems that incubation period causes differences in control-by-isolation effectivness when not at the boundary of epidemic controllability. In other words, differences between influenza subtypes arise when not in a situation of high transmissibility and low contact tracing effectivness that cannot contain any outbreaks, nor low transmissibility and high contact tracing effectivenss easily contain all outbreaks (Figure \ref{fig:prop-outbreak-control-var-R}). \\

@@

@@

The pattern of outbreak control is broadly similar between 10\% asymptomatic cases and the 0\% and 30\% asymptomatic scenarios (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). Qualitative difference are in the 30\% asymptomatic scenario for LFT isolation, where the ability to control highly transmissible influenzas reduces to less than a quarter of outbreaks controlled at maximum contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). When every case is symptomatic the likelihood of containing highly transmissible outbreaks drastically increases for all onset-to-isolation response delays (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}). \\

The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak. The higher the number of cases the less chance that stochastic extinction of a transmission chain will end the whole epidemic. For an influenza pandemic, the number of initial infected individuals influences efficacy of control-by-isolation at all transmissibility scenarios evaluated ($R \in \{1.1, 1.5, 2.5, 3.5\}$, Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.1$ most outbreaks can be controlled without isolation and contact tracing, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}). When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}). When the number of cases seeding the outbreak is $\geq 20$ at $R = 2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $\geq 80\%$. In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks; except for H1N1 where only 25\% of outbreaks are controlled if seeded with 40 cases (Figure \ref{fig:prop-outbreak-control-num-init-cases}).  \\

@@
